Substructure Boosts to Dark Matter Annihilation from Sommerfeld Enhancement

### 2009-02-22 tex/somm.tex
Wrote the conclusion

Of course, when considering the Sommerfeld enhancement of a halo of a
certain velocity dispersion $\sigv$ we need to average the Sommerfeld
enhancement the distribution of relative velocities of that halo. We
will approximate the one-particle velocity distribution of the halo as
a single truncated Maxwell-Boltzmann distribution with velocity
dispersion $\sigv$
\begin{equation}\label{eq:maxboltz}
f(v) \propto \left\{
\begin{array}{rl}
v^2\, e^{-v^2/2\sigv^2}  & \qquad v \leq v_{\mathrm{esc}}\\
0 & \qquad v > v_{\mathrm{esc}}\, ,\\
\end{array} \right.
\end{equation}
in which we set the escape velocity to $v_{\mathrm{esc}} =
\sqrt{2}\,\sigv$. Note that in the pure $1/v$ regime this averaging
will tend to sustain the $1/v$ behavior up to a factor $\order(1)$,
as we have that
\begin{equation}
\int_0^\infty \dd v \, v^2\, e^{-v^2/4\sigv^2}
\somm\left/\int_0^\infty \dd v \, v^2\, e^{-v^2/4\sigv^2} \right.\propto
\frac{1}{\sigv}\, ,
\end{equation}
since the truncation does not significantly influence the
result. Similarly, in the resonance region the $1/v^2$ behavior is
conserved by the averaging. When the enhancement has leveled off the
averaging has no effect.

@@

\begin{figure*}[t]
\includegraphics[clip=]{flux_carina_65.eps}%%BoundingBox: 178 291 400 474
\includegraphics[clip=]{flux_draco_65.eps}\\%%BoundingBox: 190 291 430 474
\includegraphics[clip=]{flux_fornax_65.eps}%%BoundingBox: 178 291 400 474
\includegraphics[clip=]{flux_leoI_65.eps}\\%%BoundingBox: 190 291 430 474
\includegraphics[clip=]{flux_leoII_65.eps}%%BoundingBox: 178 291 400 474
\includegraphics[clip=]{flux_ursaminor_65.eps}%%BoundingBox: 190 291 430 474
\caption{}%
\label{fig:dsph}%
\end{figure*}

@@

\section{Predicted annihilation signals for dwarf Spheroidals}

@@

\begin{figure}
\centering
\includegraphics[clip=]{hist_-2_-3_10_1.eps}
\caption{The distances to and masses of subhalos within the tidal
radius of the main halo in \VL\ II. The distances are for three random
observers at a distance of 8.5 kpc from the galactic center of the
main halo.}%\
\label{fig:VL23_1}%
\end{figure}

@@

\section{Discussion}

Don't forget to mention the fact that the whole halo might be build up
of distinct streams with velocity dispersion of order 1 km/s, such
that the annihilation signal we might expect from the halo could be
much larger than that people have suggested.

>>>>

Of course, when considering the Sommerfeld enhancement of a halo of a
certain velocity dispersion $\sigv$ we need to average the Sommerfeld
enhancement the distribution of relative velocities of that halo. We
will approximate the one-particle velocity distribution of the halo as
a single truncated Maxwell-Boltzmann distribution with velocity
dispersion $\sigv$
\begin{equation}\label{eq:maxboltz}
f(v) \propto \left\{
\begin{array}{rl}
v^2\, e^{-v^2/2\sigv^2}  & \qquad v \leq v_{\mathrm{esc}}\\
0 & \qquad v > v_{\mathrm{esc}}\, ,\\
\end{array} \right.
\end{equation}
in which we set the escape velocity to $v_{\mathrm{esc}} =
\sqrt{2}\,\sigv$. This approximation of a single Maxwell-Boltzmann
distribution for the velocity distribution means that we assume that
the halo has a constant velocity dispersion over the halo, \ie, that
the halo is isothermal. This is not necessarily the case ---indeed, it
is not if we believe that the density profiles of \DM\ halos are well
fit by Einasto or \NFW\ profiles--- but it is not a bad assumption for
intermediate distances from the halo's center, \ie, between the
innermost and outermost parts of the halo
\cite{2007ApJ...667L..53W}. A more detailed treatment of the
Sommerfeld enhancement in subhalos should take the non-uniform
velocity dispersion in subhalos into account. Note that in the pure
$1/v$ regime this averaging will tend to sustain the $1/v$ behavior up
to a factor $\order(1)$, as we have that
\begin{equation}
\int_0^\infty \dd v \, v^2\, e^{-v^2/4\sigv^2}
\somm\left/\int_0^\infty \dd v \, v^2\, e^{-v^2/4\sigv^2} \right.\propto
\frac{1}{\sigv}\, ,
\end{equation}
since the truncation does not significantly influence the
result. Similarly, in the resonance region the $1/v^2$ behavior is
conserved by the averaging. When the enhancement has leveled off the
averaging has no effect.

@@

\begin{figure*}[t]
\includegraphics[clip=]{flux_carina_65.eps}%%BoundingBox: 178 291 400 474
\includegraphics[clip=]{flux_draco_65.eps}\\%%BoundingBox: 190 291 430 474
\includegraphics[clip=]{flux_fornax_65.eps}%%BoundingBox: 178 291 400 474
\includegraphics[clip=]{flux_leoI_65.eps}\\%%BoundingBox: 190 291 430 474
\includegraphics[clip=]{flux_leoII_65.eps}%%BoundingBox: 178 291 400 474
\includegraphics[clip=]{flux_ursaminor_65.eps}%%BoundingBox: 190 291 430 474
\caption{Predicted \DM\ annihilation fluxes for various dwarf
Spheroidal satellite galaxies. The properties of the \dSphs\ used are
given in table \ref{tab:dsphs}.}%
\label{fig:dsph}%
\end{figure*}

@@

\section{Predicted annihilation signals for dwarf Spheroidals}\label{sec:dsphs}

@@

\begin{figure}
\centering
\includegraphics{hist_-2_-3_10_1.eps}
\caption{The distances to and masses of subhalos within the tidal
radius of the main halo in \VL\ II. The distances are for three random
observers at a distance of 8.5 kpc from the galactic center of the
main halo.}%\
\label{fig:VL23_1}%
\end{figure}

@@

\section{Conclusion}

The recently proposed Sommerfeld enhancement to the \DM\ annihilation
cross section forces us to reconsider statements made in the past
about the expected signal of \DM\ anihilation from cold substructures
in the halo of the Galaxy. Since the Sommerfeld enhancement depends
criticically on the velocity dispersion of the substructure in
question, as well as on the details of the particle physics model used
to describe the attractive (or, possibly, repulsive) force that gives
rise to the Sommerfeld enhancement, the contributions from the
specific particle physics model and the details of the \DM\
distribution in the substructure can no longer be as cleanly separated
as they had been before. This means that in a scenario in which there
is a significant Sommerfeld enhancement, previous claims of ``particle
physics independent'' conclusions about the \DM\ annihilation signal
from substructure (\eg,
\cite{2007PhRvD..75h3526S,2008Natur.456...73S}) need to be
reconsidered in this new framework. This complicates the analysis but
it opens up new possibilities for detecting \DM\ annihilation and
learning about the properties of the dark sector. Given a detailed
knowledge of the density structure and velocity dispersion profile of
the dwarf Spheroidal galaxies, much can be learned if we are able to
detect the \DM\ annihilation signal coming from these \dSphs, by
considering the relative signal and using the results of
Sec.~\ref{sec:dsphs}, or by using the detailed velocity dispersion
profile of the \dSphs\ \cite{2009arXiv0902.0362R}.

We have focused on the consequences of the Sommerfeld enhancement for
\DM\ subhalos in a \LCDM\ cosmology, using the latest results from
high-resolution numerical $N$-body simulations to inform our
assumptions about the abundance and properties of the
substructure. Under these assumptions we were able to show that there
are very large boosts to the \DM\ annihilation cross section from the
combined effect of substructure boosts and Sommerfeld enhancement in
large parts of the parameter space consisting of the parameters of the
attractive Yukawa type force between the \DM\ particles which is
responsible for the Sommerfeld enhancement. These regions of parameter
space are in no way unrealistic. Indeed, in one of the most
straightforward interpretations of the ATIC/PAMELA results the sharp
cut-off around 1 TeV seen in the ATIC spectrum is interpreted as being
due to the fact that the dark matter mass is about 700 GeV. The
absence of an excess in anti-protons similar to the excess in
positrons seen by PAMELA can be explained by positing that the dark
matter particles annihilate to an intermediate boson $\phi$, which is
lighter than a proton (or even, lighter than the effective quark mass
$\Lambda_{\mbox{\footnotesize QCD}}$) in order to kinematically forbid
the annihilation into protons \cite{2008arXiv0811.3641C}. This
naturally leads to a \mv/\mdm $\lesssim$ $10^{-3}$. The coupling
constant of the new force in the dark sector is also generally $\alpha
\gtrsim 10^{-2}$ \cite{ArkaniHamed:2008qn,2008arXiv0812.0360L}. Thus,
from Figure \ref{fig:subboost}, we expect there to be a rather large
boost to the \DM\ annihilation cross section in these models. This
leads to particularly large \DM\ annihilation signals predicted for
the \dSphs\ (see Figure \ref{fig:dsph}).

We have not discussed the substructure boost to the \DM\ annihilation
signal from unresolved substructure in the Galactic halo. This is
however, interesting in the light of the Sommerfeld enhancement, since
it might lead to a different prediction about the relative magnitudes
of the annihilation signal coming from the resolved Galactic halo and
unresolved halo, as well from the resolved substructures in the
Galactic halos. It is already clear that in the generic 1/$v$
Sommerfeld enhancement the main halo, with its typical velocity
dispersion of $\sim\!150$ km s$^{-1}$, will receive a much smaller
enhancement than the \dSphs, with their typical velocity dispersions
of 10 km s$^{-1}$. This, however, ignores the effects of saturation
and resonances, which might be important in the specific particle
physics model adopted for the Sommerfeld enhancement. Additionally,
while the velocity dispersion of the Galactic halo might be of the
order of 100 km s$^{-1}$ the actual velocity dispersion relevant to
the Sommerfeld enhancement might be much smaller for halo \DM\
particles. If a large fraction of the \DM\ in the Galactic halo is
part of a cold stream ---and various numerical simulations predict
that large parts of the halo, if not all of the halo, is made up of
such cold streams with typical velocity dispersions of the order of 1
km s$^{-1}$ \cite{1999MNRAS.307..495H,2003MNRAS.339..834H}--- then the
relevant relative velocity of \DM\ particles is set by the velocity
dispersion of particles in the stream, and not by the velocity
dispersion of the streams in the halo, which is what the 100 km
s$^{-1}$ Galactic halo velocity dispersion corresponds to in these
scenarios. Thus, the expected \DM\ annihilation signal from the main
halo could be much larger than that naively expected for a halo with a
velocity dispersion of a 100 km s$^{-1}$.
